\tit

### 2021-06-02 response-to-reviews.md

5. Fix typos and other issues that have already been discussed in the reviews and the response.
</blockquote>

My response:

1.  I added the following paragraph to the beginning of section 7 (Related Work):

    > The shift from languages as sets of strings to type-level predicates (i.e., proof relevance or “parsing”) is akin to weighted automata [Schützenberger 1961; Droste and Kuske 2019] and more generally to semiring-based parsing [Chomsky and Schützenberger 1959; Goodman 1998, 1999; Liu 2004], noting that types (“Set” in Agda) form a (commutative) semiring (up to isomorphism). In particular, Lombardy and Sakarovitch [2005] investigated Brzozowski-style derivatives in this more general setting, formalizing and generalizing the work of Antimirov [1996], who decomposed derivatives into simpler components (“partial derivatives”) that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.

@@

    Also an initial sentence to the (now) third paragraph of that section:

    > The main contributions of this paper are to formalization and mechanization of language recognition and parsing.

2.  I added a new figure 2 with examples of languages and membership proofs, along with a new footnote 2 giving some detailed explanations.

3.  I added a new footnote 10:

    > Inductive types (“data”) describe finitely large values and are processed recursively (as least fixed points) by pattern- matching clauses that decompose arguments. Dually, coinductive types (“codata”) describe infinitely large values and are processed corecursively (as greatest fixed points) by copattern-matching clauses that compose results. Programs on inductive types are often proved by induction, while programs on coinductive types are often proved by coinduction. (As we will see in Section 6.2, however, the simple relationship to the specification in Section 2 allows many trivial correctness proofs, needing neither induction or coinduction.) See Gordon [1995] for a tutorial on theory and techniques of coinduction.
    > 
    > Tries in general represent functions, with each trie datum corresponding to a domain value. Even when each domain value is finitely large, there are often infinitely many of them (e.g., lists), so tries will naturally be infinite and thus more amenable to coinductive than inductive analysis.

@@

4.  Appended to the final paragraph of section 7:

    > As mentioned at the end of Section 3, one can simplify the development above somewhat by replacing proof isomorphism with the coarser notion of logical equivalence. Doing so restores union idempotence (modulo equivalence), losing proof relevance and thus re- placing parsing by recognition. One might then also prove that the set of derivatives of a regular language is finite (again, modulo equivalence).

### 2021-06-02 paper.tex
chnages in respose to reviews

The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a trivial semantic function {\AF{⟦\_⟧}} (mapping any representation of type {\AF{Lang} \AB P} to the corresponding type-level language \AB{P}) and a less trivial, computable function  {\AF{⟦\_⟧‽}} that converts a symbolic language representation to a computable parser.
Note that the \AF{\_◂\_} operation from \secref{Decidability} has been reified as a new constructor alongside the more conventional language operators.
The reason \AF{\_◂\_} must be part of the inductive representation is the same as the other constructors, namely so that the core lemmas (\figref{nu-delta-lemmas}) translate into an implementation in terms of that representation.

@@

\secref{Symbolic Differentiation} embodies one choice of reflecting functions into a data representation.
Now consider the dual strategy of ``automatic differentiation''.
This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.\footnote{Inductive types (``data'') describe finitely large values and are processed recursively (as least fixed points) by pattern-matching clauses that decompose arguments.
Dually, coinductive types (``codata'') describe infinitely large values and are processed recursively (as greatest fixed points) by copattern-matching clauses that compose results.
Programs on inductive types are often proved by induction, while programs on coinductive types are often proved by coinduction\out{ and specifically bisumulation}.
(As we will see in \secref{Transporting Properties from Specification to Implementations}, however, the simple relationship to the specification in \secref{Decomposing Languages} allows many trivial correctness proofs, needing neither introduction or coinduction.)
See \citet{Gordon1995:coinduction} for a tutorial on theory and techniques of coinduction.

Tries in general represent functions, with each trie datum corresponding to domain value.
Even when each domain value is finitely large, there are often infinitely many of them (e.g., lists), so tries will naturally be infinite and thus more amenable to coinductive than inductive analysis.}
(This program duality pattern has been noted and investigated by \citet{OstermannJabs2018}.)
Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\api}.

@@

The shift from languages as sets of strings to type-level predicates (i.e., proof relevance or ``parsing'') is akin to weighted automata \citep{Schutzenberger1961, DrosteKuske2019} and more generally to semiring-based parsing \citep{Chomsky1959CFL, Goodman1998PIO, Goodman1999SP, Liu2004}, noting that types (``\APT{Set}'' in Agda) form a (commutative) semiring (up to isomorphism).
In particular, \citet{Lombardy2005} investigated Brzozowski-style derivatives is this more general setting, formalizing and generalizing the work of \citet{Antimirov1996}, who decomposed derivatives into simpler components (``partial derivatives'') that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.

>>>>

The ``symbolic differentiation'' style reflects language-building vocabulary (\figref{Lang-ops}) into an inductive data type (of modestly extended regular expressions), while \AF{ν} and \AF{δ} become functions defined by pattern-matching on that data type.
For convenience, use the same names as in \secref{Specifying Languages} for these new counterparts, while referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
The result is shown in \figrefdef{symbolic-api}{Regular expressions (inductive)}{\api} along with a trivially defined semantic function {\AF{⟦\_⟧}} (mapping any representation of type {\AF{Lang} \AB P} to the corresponding type-level language \AB{P}) and a less trivial, computable function  {\AF{⟦\_⟧‽}} that converts a symbolic language representation to a computable parser.
Note that the \AF{\_◂\_} operation from \secref{Decidability} has been reified as a new constructor alongside the more conventional language operators.
The reason \AF{\_◂\_} must be part of the inductive representation is the same as the other constructors, namely so that the core lemmas (\figref{nu-delta-lemmas}) translate into an implementation in terms of that representation.

@@

\secref{Symbolic Differentiation} embodies one choice of reflecting functions into a data representation.
Now consider the dual strategy of ``automatic differentiation''.
This time, reflect \AF{ν} and \AF{δ} into a \emph{coinductive} data type of tries\footnote{The classic trie (``prefix tree'') data structure was introduced by \citet{Thue1912Gegenseitige} to represent sets of strings \citep[Section 6.3]{Knuth1998ACP3}, later generalized to arbitrary non-nested algebraic data types \citep{Connelly1995GenTrie}, and then from sets to functions \citep{Hinze2000GGT}.}, while redefining the language-building vocabulary as functions on that data type defined by \emph{copatterns} \citep{AbelPientka2016}.\footnote{Inductive types (``data'') describe finitely large values and are processed recursively (as least fixed points) by pattern-matching clauses that decompose arguments.
Dually, coinductive types (``codata'') describe infinitely large values and are processed corecursively (as greatest fixed points) by copattern-matching clauses that compose results.
Programs on inductive types are often proved by induction, while programs on coinductive types are often proved by coinduction\out{ and specifically bisumulation}.
(As we will see in \secref{Transporting Properties from Specification to Implementations}, however, the simple relationship to the specification in \secref{Decomposing Languages} allows many trivial correctness proofs, needing neither induction or coinduction.)
See \citet{Gordon1995:coinduction} for a tutorial on theory and techniques of coinduction.

Tries in general represent functions, with each trie datum corresponding to a domain value.
Even when each domain value is finitely large, there are often infinitely many of them (e.g., lists), so tries will naturally be infinite and thus more amenable to coinductive than inductive analysis.}
(This program duality pattern has been noted and investigated by \citet{OstermannJabs2018}.)
Again, we will use the same names as in \secref{Specifying Languages}, referring to the original versions via the module prefix ``{◇.}\hspace{0.05em}''.
The result is shown in \figrefdef{automatic-api}{Tries (coinductive)}{\api}.

@@

The shift from languages as sets of strings to type-level predicates (i.e., proof relevance or ``parsing'') is akin to weighted automata \citep{Schutzenberger1961, DrosteKuske2019} and more generally to semiring-based parsing \citep{Chomsky1959CFL, Goodman1998PIO, Goodman1999SP, Liu2004}, noting that types (``\APT{Set}'' in Agda) form a (commutative) semiring (up to isomorphism).
In particular, \citet{Lombardy2005} investigated Brzozowski-style derivatives in this more general setting, formalizing and generalizing the work of \citet{Antimirov1996}, who decomposed derivatives into simpler components (``partial derivatives'') that sum to the full derivative and lead to efficient recognizers in the form of nondeterministic automata.
